\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-03-11 docs/parco13/abstract.tex
Some text on parallelization

\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}\xspace}
\newcommand{\sR}{\mathbb{R}}
\newcommand{\forestclaw}{Forestclaw\xspace}
\newcommand{\pforest}{\texttt{p4est}\xspace}

@@

In this paper we describe how we design the coupling of forest-of-octree AMR
with Clawpack at the leaf level.  We comment on challenges that arise in
enabling efficient parallelism and multiblock geometries and close with a range
of numerical examples that demonstrate the conceptual improvements in relation
to other approaches.

@@

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the Clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{Clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
structure & forest of octrees & patch of $m^d$ DOFs \\
atomic unit & octree leaf & one DOF \\
parallelization & MPI & threads (Manyclaw variant) \\
memory access & distributed & shared \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
dependencies & none & Python \\
%\hline
\end{tabular}
\end{center}
The proposed 1:1 correspondence between a leaf and a patch thus combines two
hithertwo disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified and performant codes.
% if the integration is done well
\item We preserve the separation of the mesh on one hand from the
discretization and solvers on the other.
\item The AMR metadata (24 bytes per leaf) is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid and often referred
to as MPI+X.
\end{enumerate}

@@

A main challenge is presented by the fact that the patch neighborhood is only
known to the mesh backend, from which it needs to be propagated to the
numerical code in \forestclaw that implements the interaction with ghost
patches.  To this end, we define an interface that allows distributed read-only
access to the sequence of blocks, the list of patches for each, and a lookup of
neighbor patches (and their relative orientation which is nontrivial between
blocks).  The exchange of ghost data and parallel partitioning is provided
transparently by the backend.  Mesh modification directives, such as local
refinement and coarsening, are called from \forestclaw and relayed to the mesh
backend.

@@

$z$-order

Iterators over all leafs, optionally restricted to a given level.
Random access is possible.  Looping over the patches in the order prescribed
by the forest leads to a high percentage of cache reuse, since the ghost
patches are likely closeby with respect to the $z$-order.

same block, same rank: local copy.

neighbor block: permutation of patch points due to non-aligned
coordinate systems

neighbor rank: data needs to be fetched over the network first.

Trees have different coordinate systems.

$L$ is 30 for 2D and 19 for 3D, so it allows for deep hierarchies.

fast (O(log N/P)) neighbor and parent/child lookups
horizontal/vertical tree traversal

subcycling: level-wise load balancing as done in structured AMR or
geometric adaptive multigrid \cite{SundarBirosBursteddeEtAl12}

>>>>

\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}\xspace}
\newcommand{\sR}{\mathbb{R}}
\newcommand{\dt}{\mathrm{d}t}
\newcommand{\forestclaw}{Forestclaw\xspace}
\newcommand{\pforest}{\texttt{p4est}\xspace}

@@

In this paper we describe how we design the coupling of forest-of-octree AMR
with Clawpack at the leaf level.  We comment on challenges that arise in
enabling multiblock geometries and efficient parallelism and close with a range
of numerical examples that demonstrate the conceptual improvements in relation
to other approaches.

@@

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the Clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{Clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
structure & forest of octrees & patch of $m^d$ DOFs \\
atomic unit & octree leaf & one DOF \\
parallelization & MPI & threads (Manyclaw variant) \\
memory access & distributed & shared \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
dependencies & none & Python \\
%\hline
\end{tabular}
\end{center}
The proposed 1:1 correspondence between a leaf and a patch thus combines two
hithertwo disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified and performant codes.
% if the integration is done well
\item We preserve the separation of the mesh on one hand from the
discretization and solvers on the other.
\item The AMR metadata (24 bytes per leaf) is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).
\end{enumerate}

@@

A main challenge is presented by the fact that the patch neighborhood is only
known to \pforest, from which it needs to be propagated to the
numerical code in \forestclaw that implements the interaction with ghost
patches.  To this end, we define an interface that allows distributed read-only
access to the sequence of blocks, the list of patches for each, and a lookup of
neighbor patches (and their relative orientation which is nontrivial between
blocks).  The exchange of ghost data and parallel partitioning is provided
transparently by \pforest.  Mesh modification directives, such as local
refinement and coarsening, are called from \forestclaw and relayed to \pforest.

% Forestclaw would support swapping out p4est or Clawpack

@@

The MPI layer is addressed from within \pforest and not exposed to the
\forestclaw code.  The order of leaves is maintained in \pforest according to a
space filling curve.  Each MPI rank has a local view on its own partition,
augmented where necessary with information about one layer of ghost patches.

\forestclaw uses iterators over all leafs, optionally restricted to a given
level.  Random access is possible and used when executing $O(1)$ time neighbor
lookups.  Looping over the patches in the order prescribed by the forest and
accessing neighbors only relative to the current patch leads to a high
percentage of cache reuse.
% since the ghost patches are likely closeby with respect to the $z$-order.

When \forestclaw accesses neighbor patches, they can be on the same or a
different block.  In the latter case, coordinate transformations are carried
out.  The structure of \forestclaw is oblivious to the fact that it only has a
local view of the distributed mesh and data which relieves the developer from
programming to the MPI interface.  Parallel neighbor exchanges are hidden
inside \pforest and called by \forestclaw at well-defined synchronization
points.  There is one such parallel data exchange per time step for a global
value of $\dt$, or one exchange per discretization level per time step if $\dt$
is chosen per-patch depending on its size (this is sometimes called
subcycling).  When using subcycling, the load balance is attained per level as
it is done in structured AMR or recent geometric adaptive multigrid schemes
\cite{SundarBirosBursteddeEtAl12}.

The threaded parallelism over the degrees of freedom of a patch is handled by
\forestclaw alone without the need to involve \pforest.

% Trees have different coordinate systems.
% neighbor block: permutation of patch points due to non-aligned coordinate
% systems

% neighbor rank: data needs to be fetched over the network first.

% $L$ is 30 for 2D and 19 for 3D, so it allows for deep hierarchies.

% fast (O(log N/P)) neighbor and parent/child lookups
% horizontal/vertical tree traversal
